---
title: "An observational analysis of the effect of improved sanitation on linear growth in the WASH Benefits Bangladesh and Kenya trials"
author: ""
date: "13 Nov 2017 (updated 28 Mar 2018)"
output: 
  html_notebook:
    theme: default
    highlight: haddock
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Overview
This script conducts an observational analysis of the relationship between improved sanitation at enrollment and LAZ at the final endpoint in the WASH Benefits trials. This is a prospective cohort study nested within the control arm of each trial, whereby our exposure is household sanitation conditions at enrollment and our outcome is length-for-age z score (LAZ) at the final endpoint, when children were about 2 years old (median age in Kenya = 22 months, median age in Bangladesh = 25 months).  

In the Bangladesh trial, we defined an improved latrine as a latrine with a functional water seal. In the Kenya trial, we followed the standard WHO JMP definition for access to an improved latrine.  These definitions reflect locally appropriate definitions of improved sanitation, as used in reporting the original trials.

We restricted the analyses to the control arms of each trial to avoid any complexity due to effects of intervention (in the case of nutritional arms) or changes in sanitation conditions as a result of the trial (in the case of sanitation arms).

The section Summary of Results includes the results across both countries, so if you want to skip the details and cut to the chase then click on that section in the TOC (at left) to go to the end.

To download the R markdown source code for this notebook, toggle the dropdown menu at the top right of this page to "Download Rmd".

## Preamble
```{r preamble, eval=TRUE,echo=TRUE,results='hide'}
# Preamble - load packages and save colors for plotting
# WASH Benefits trial analysis package
# dev. version at devtools::install_github("ben-arnold/washb")
library(washb) 
# Machine learning and TMLE packages
library(SuperLearner)
library(tmle)
library(sandwich)
library(lmtest)
library(zoo)

library(gam)
library(splines)
library(glmnet)
library(arm)
library(foreach)
# dm and data visualization
library(tidyverse)

# bright color blind palette:  https://personal.sron.nl/~pault/ 
cblack <- "#000004FF"
cblue <- "#3366AA"
cteal <- "#11AA99"
cgreen <- "#66AA55"
cchartr <- "#CCCC55"
cmagent <- "#992288"
cred <- "#EE3333"
corange <- "#EEA722"
cyellow <- "#FFEE33"
cgrey <- "#777777"
```


# Bangladesh analysis

## Load and process the Bangladesh data

Note: these datasets are publicly available through the Open Science Framework:
https://osf.io/pqzj5

```{r load bangladesh data, results='hide',eval=TRUE}
# load and merge the final analysis files
# treatment assignments, enrollment charactersitics, and anthropometry measurements
d_tr    <- read.csv("~/washb-obs-san/data/washb-bangladesh-tr-public.csv")
d_enrol <- read.csv('~/washb-obs-san/data/washb-bangladesh-enrol-public.csv')
d_anth  <- read.csv('~/washb-obs-san/data/washb-bangladesh-anthro-public.csv')

#  merge to final analysis files, loaded above
lazd <- left_join(d_enrol,d_tr,by=c("clusterid","block"))
lazd <- left_join(lazd,d_anth,by=c("dataid","clusterid","block"))

# filter to children with measurements at year 2 with valid LAZ measuresments
# filter to the control arm
d2 <- lazd %>%
  filter(svy==2 & laz_x==0 & tr %in% c("Control"))

# simplify the dataset by selecting covariates and exposures
# these are the pre-specified covariates from the trial
Ws <- c("month","aged","sex","birthord","momage","momedu","momheight","Nlt18","Ncomp","watmin","hfiacat","elec","floor","walls","roof","asset_wardrobe","asset_table","asset_chair","asset_khat","asset_chouki","asset_tv","asset_refrig","asset_bike","asset_moto","asset_sewmach","asset_mobile")

d <- d2 %>%
  select(block,laz,latown,latseal,Ws) %>%
  mutate(latrine=factor(ifelse(latown==0 | is.na(latown),"No latrine","Latrine")),
         implatrine=factor(ifelse(latown==1 & latseal==1,"Improved latrine","No improved latrine")))

d$implatrine[is.na(d$implatrine)] <- "No improved latrine"

d$latcat <- factor(NA,levels=c("No latrine","Latrine no water seal","Latrine with water seal"))
d$latcat[d$latrine=="No latrine"] <- "No latrine"
d$latcat[d$latrine=="Latrine" & d$implatrine=="No improved latrine"] <- "Latrine no water seal"
d$latcat[d$latrine=="Latrine" & d$implatrine=="Improved latrine"] <- "Latrine with water seal"


```

## Distribution of LAZ at the two year measurement by access to improved latrines

This is a summary of children in the control arm at the final endpoint, stratified by latrine access at enrollment:
```{r bangladesh table}
# table of latrine type, with mean LAZ and SDs
b_lattable <- as.data.frame(table(d$latcat))

# mean and SD
b_musd <- d %>%
  group_by(latcat) %>%
  select(latcat, laz) %>%
  summarize(n=n(),
            pct=sprintf("%1.0f",100*(n/sum(b_lattable$Freq))),
            mean=mean(laz),
            sd=sd(laz))



colnames(b_musd) <- c("Sanitation at enrollment","N children","(%)","Mean LAZ","SD LAZ")
knitr::kable(b_musd,digits=c(0,0,0,3,3),format="markdown")
```

This is a summary of the LAZ distributions by the different groups:
```{r bangladesh density}

# plot distributions of LAZ stratified by latrine ownership at enrollment
p1 <- ggplot(data=filter(d,latcat!="Latrine with water seal"), aes(x=laz,group=latcat,fill=latcat)) +
  geom_density(alpha=0.4) +
  scale_fill_manual(values=c(corange,cmagent),
                    guide=guide_legend(title="Sanitation at enrollment")) +
  theme_minimal()
p1

p2 <- ggplot(data=filter(d,latcat!="Latrine no water seal"), aes(x=laz,group=latcat,fill=latcat)) +
  geom_density(alpha=0.4) +
  scale_fill_manual(values=c(corange,cteal),
                    guide=guide_legend(title="Sanitation at enrollment")) +
  theme_minimal()
p2


```

## Association between access to an improved latrine and LAZ

Estimate the difference in LAZ between children born into households that have a latrine with a functional water seal versus those born into households without a latrine. The estimator is a genearlized linear model (GLM) with robust, sandwich standard errors clustered at the block level, which was the independent unit in the trial. Adjusted estimates screen pre-specified covariates using a likelihood ratio test. 
```{r ests}
# unadjusted
b_glmest_unadj <- washb_glm(Y=d$laz,tr=d$latcat,id=d$block,
                            contrast=c("No latrine","Latrine with water seal"))

# adjusted
b_glmest_adj <- washb_glm(Y=d$laz,tr=d$latcat,W=d[Ws],id=d$block,
                          contrast=c("No latrine","Latrine with water seal"))

```

## Double-robust, targeted maximum likelihood estimation (TMLE)

This estimator follows the pre-specified, adjusted analysis methods from the main trial. In our view, this is the most aggressive statistical adjustment possible given the design of a fixed exposure and prospectivly measured outcome. Targeted maximum likelihood estimation is a double-robust approach that models both the outcome and the treatment mechanism (here: access to improved sanitation at enrollment). It allows for flexible estimation of each model using ensemble machine learning. We used the same ensemble and pre-specified covariates as was used in the trial's primary analysis (simple mean, GLM, Bayes GLM, GAM with cubic splines, LASSO). Standard errors are based on the influence curve and treat blocks as the independent unit (as in the main trial).

```{r tmleest}
set.seed(12345)
b_tmleest_adj <- washb_tmle(Y=d$laz,tr=d$latcat,W=d[Ws],
                          id=d$block,
                          contrast=c("No latrine","Latrine with water seal"),
                          Q.SL.library=c("SL.mean","SL.glm","SL.bayesglm","SL.gam","SL.glmnet"))


```



# Kenya analysis

## Load and process the Kenya data

Note: these datasets are publicly available through the Open Science Framework:
https://osf.io/uept9/
```{r load kenya data, results='hide',eval=TRUE}

# load and merge the final analysis files
# In Kenya, the anthropometry data from visit 2 already has
# covariates merged to it (including treatment information)
# However, one of the pre-specified covariates was not
# improved latrine ownership at enrollment, so add that from
# the data used to create Table 1 from the manuscript
dk_enrol <- read.csv("~/washb-obs-san/data/washb-kenya-uptake-baseline-public.csv")
dk_tr    <- read.csv("~/washb-obs-san/data/washb-kenya-tr-public.csv")
dk_anth  <- read.csv("~/washb-obs-san/data/washb-kenya-endline-anthro-public.csv")

# restrict enrollment measures to improved latrine and merge
# to the anthropometry measures at the year 2 visit
dk_enrol <- dk_enrol %>%
  select(clusterid,hhid,implat=imp_lat)
dk <- left_join(dk_tr,dk_enrol,by=c("clusterid"))
dk <- left_join(dk,dk_anth,by=c("block","clusterid","hhid"))

# filter to just index children in the control arms
dk <- dk %>%
  filter(targetchild==1 & tr %in% c("Control","Passive Control"))

# there were 150/2272 children (6%) for whom improved latrine status 
# was missing.  Exclude them (missing at random assumption)
# there were 21 with unrealistic LAZ measurements. Exclude them
dk <- dk %>% 
  filter(!is.na(implat) & !is.na(haz))

# select pre-specified covariates from the primary outcomes
# excluding staffid since it is sparse without extensive recoding
# ensure factors are coded correctly
# this follows the primary outcome adjusted analysis (line ~74):
# https://github.com/jadebc-berkeley/WBK-primary-outcomes/blob/master/primary/analysis/0-base-programs.R
kWs <- c("month","HHS","agey","sex","mother_age","motherht","mother_edu",
    "u18","Ncomp","water_time","floor","roof","cow",
    "goat","chicken","dog","elec","radio","tv","mobilephone",
    "clock","bicycle","motorcycle","stove")

kd <- dk %>%
  select(block,clusterid,tr,laz=haz,implat,kWs)

# reorder factor values to so that the reference is 
# not having something instead of the first thing alphabetically
kd$month <-  as.factor(kd$month)
kd$clock <- relevel(kd$clock,ref="No clock")
kd$elec <- relevel(kd$elec,ref="No electricity")
kd$mother_edu=relevel(kd$mother_edu,ref="Incomplete primary")
kd$roof <- relevel(kd$roof,ref="Thatch/leaf")
kd$floor <- relevel(kd$floor,ref="Earth/dung")
kd$bicycle <- relevel(kd$bicycle,ref="No bicycle")
kd$radio <- relevel(kd$radio,ref="No radio")
kd$tv <- relevel(kd$tv,ref="No TV")
kd$mobilephone <- relevel(kd$mobilephone,ref="No mobile phone")
kd$stove <- relevel(kd$stove,ref="No stove")

# formally code improved latrine
kd$latcat <- factor(ifelse(kd$implat==1,"Improved latrine","No improved latrine"),levels=c("No improved latrine","Improved latrine"))

```


## Distribution of LAZ at the two year measurement by access to improved latrines

This is a summary of children in the active and passive control arms at the final endpoint, stratified by latrine access at enrollment:
```{r kenya table}
# table of latrine type, with mean LAZ and SDs
k_lattable <- as.data.frame(table(kd$latcat))

# mean and SD
k_musd <- kd %>%
  group_by(latcat) %>%
  select(latcat, laz) %>%
  summarize(n=n(),
            pct=sprintf("%1.0f",100*(n/sum(k_lattable$Freq))),
            mean=mean(laz),
            sd=sd(laz))

colnames(k_musd) <- c("Sanitation at enrollment","N children","(%)","Mean LAZ","SD LAZ")
knitr::kable(k_musd,digits=c(0,0,0,3,3),format="markdown")
```

This is a summary of the LAZ distributions by the different groups:
```{r kenya density}
# plot distributions of LAZ stratified by baseline latrine ownership
ggplot(data=kd,aes(x=laz,group=latcat,fill=latcat)) +
  geom_density(alpha=0.4) +
  scale_fill_manual(values=c(corange,cteal),
                    guide=guide_legend(title="Sanitation at enrollment")) +
  theme_minimal()

```

## Association between access to an improved latrine and LAZ

Estimate the difference in LAZ between children born into households that have a latrine with a functional water seal versus those born into households without a latrine. The estimator is a genearlized linear model (GLM) with robust, sandwich standard errors clustered at the block level, which was the independent unit in the trial. Adjusted estimates screen pre-specified covariates using a likelihood ratio test. 

```{r kenya ests}
# unadjusted
k_glmest_unadj <- washb_glm(Y=kd$laz,tr=kd$latcat,id=kd$block,
                            contrast=c("No improved latrine","Improved latrine"))

# adjusted
k_glmest_adj <- washb_glm(Y=kd$laz,tr=kd$latcat,W=kd[kWs],id=kd$block,
                          contrast=c("No improved latrine","Improved latrine"))

```

## Double-robust, targeted maximum likelihood estimation (TMLE)

This estimator follows the pre-specified, adjusted analysis methods from the main trial. In our view, this is the most aggressive statistical adjustment possible given the design of a fixed exposure and prospectivly measured outcome. Targeted maximum likelihood estimation is a double-robust approach that models both the outcome and the treatment mechanism (here: access to improved sanitation at enrollment). It allows for flexible estimation of each model using ensemble machine learning. We used the same ensemble and pre-specified covariates as was used in the trial's primary analysis (simple mean, GLM, Bayes GLM, GAM with cubic splines, LASSO).  Standard errors are based on the influence curve and treat blocks as the independent unit (as in the main trial).

```{r kenya tmleest}
set.seed(12345)
k_tmleest_adj <- washb_tmle(Y=kd$laz,tr=kd$latcat,W=kd[kWs],
                          id=kd$block,
                          contrast=c("No improved latrine","Improved latrine"),
                          Q.SL.library=c("SL.mean","SL.glm","SL.bayesglm","SL.gam","SL.glmnet"))


```




# Summary of results

```{r summary of results, eval=TRUE, include=FALSE}

# Bangladesh results
bests <- rbind(b_glmest_unadj$TR[c(1,2,3,6)],
               b_glmest_adj$TR[c(1,2,3,6)],
               unlist(b_tmleest_adj$estimates$ATE[c("psi","CI","pvalue")])
               )

kests <- rbind(k_glmest_unadj$TR[c(1,2,3,6)],
               k_glmest_adj$TR[c(1,2,3,6)],
               unlist(k_tmleest_adj$estimates$ATE[c("psi","CI","pvalue")])
               )
rownames(bests) <- rownames(kests) <- c("Unadjusted","Adjusted, GLM","Adjusted, double-robust TMLE")

colnames(bests) <- colnames(kests) <- c("Diff LAZ", "Lower 95%","Upper 95%", "P-value")
                                        
```

In both trials, access to improved sanitation at enrollment was associated with improved LAZ at the final endpoint. In unadjusted analyses, the effects were +```r sprintf("%1.2f",bests[1,1])```Z (Bangladesh) and +```r sprintf("%1.2f",kests[1,1])```Z (Kenya).  

Effects were attenuated after adjusting for a broad set of potentially confounding covariates using ensemble machine learning and double-robust, targeted maximum likelihood estimation. In Bangladesh, the effect was reduced to +```r sprintf("%1.2f",bests[3,1])```Z (p=```r sprintf("%1.3f",bests[3,4])```).  In Kenya, the effect was reduced to +```r sprintf("%1.2f",kests[3,1])```Z (p=```r sprintf("%1.3f",kests[3,4])```).  

Adjusted estimates using simple generalized linear models (GLM) were very similar to the more aggressivly adjusted TMLE estimates. The consistency in the point estimates between the two adjusted analyses suggests that the residual bias from unobserved confounding cannot be salvaged from more aggressive estimation approaches, even double-robust estimators that include flexible machine learning.

There was definitively no effect of improved sanitation (or any combination of WASH interventions) on LAZ in either trial when evaluating the effect of the randomized interventions.

Bangladesh trial results (see Fig 3B): https://www.ncbi.nlm.nih.gov/pubmed/29396217

Kenya trial results (see Fig 3C): https://www.ncbi.nlm.nih.gov/pubmed/29396219

The contrast between the trials' findings and these observational analysis results underscore the importance of randomized controlled trials for exposures and outcomes that are inextricably linked with broader improvements in overall wellbeing.

## Bangladesh
```{r summary table Bangladesh,echo=FALSE,include=TRUE}
knitr::kable(bests,digits=3,caption="Summary of the association between access to an improved latrine at enrollment and LAZ among the birth cohort at the final endpoint, Bangladesh")
```

## Kenya
```{r summary table Kenya,echo=FALSE,include=TRUE}
knitr::kable(kests,digits=3,caption="Summary of the association between access to an improved latrine at enrollment and LAZ among the birth cohort at the final endpoint, Kenya")
```

# Session Info
```{r session info}
sessionInfo()
```



